For induction-induction, induction is enough

### 2019-09-25 paper/universaliit.tex
small rephrasings

For a type $A$ in a context $\Gamma$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec(A)=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\Gamma}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\ \Gamma$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\Gamma\sim\model\Gamma$ and $\rec(\Gamma)=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
related to $\Gamma$. Then, the type of the relation for types is
\[
  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
  (\model\Gamma : \model\Con)
  \ra
  (\Gamma \sim \model\Gamma)
  \ra
  (\model{A}:\model\Ty\model\Gamma)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
As we do not assume that our metatheory
supports this kind of recursive-recursive definitions, we refuse to further consider
 such a possiblity, and instead just remove the 
hypothesis $\Gamma\sim\model\Gamma$ from the list of arguments.
We proceed similarly for terms and substitutions. Actually, 
this removal is not without harm. Consider for example the case of relating the
empty substitution $\Gamma\vdash \epsilonu \Rightarrow \emptyu $
with a semantic substitution $\model\sigma$ between semantic contexts
$\model\Gamma$ and $\model\Delta$. We would like to assert that $\model\sigma$
equals the empty semantic substitution $\model\epsilon$, but this is not
possible because typechecking requires that $\model\Delta$ is 
the empty semantic context. This is precisely what ensured the removed hypothesis
$\syn\cdot \sim \model\Delta$.
Our way out here is to state that $\model\sigma$ relates to the empty
substitution if the target semantic context is empty, and, acknowledging
this equality, if $\model\sigma$ is
the empty substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

@@

\subsection{Recursion}
Recursion consists in building a morphism from the syntactic model to the
semantic one.
The image of a syntactic context is going to be a semantic context which relates
to it (which is unique thanks to right uniqueness of the relation), and
similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we construct by induction on well-typed judgments the
semantic counter-parts of the syntax:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  Right uniqueness of the relation is used in this induction.

>>>>

For a type $A$ in a context $\Gamma$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec(A)=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\Gamma}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\ \Gamma$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\Gamma\sim\model\Gamma$ and $\rec(\Gamma)=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
related to $\Gamma$. Then, the type of the relation for types is
\[
  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
  (\model\Gamma : \model\Con)
  \ra
  (\Gamma \sim \model\Gamma)
  \ra
  (\model{A}:\model\Ty\model\Gamma)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
As we do not assume that our metatheory
supports this kind of recursive-recursive definitions, we refuse to further consider
 such a possiblity, and instead just remove the 
hypothesis $\Gamma\sim\model\Gamma$ from the list of arguments.
We proceed similarly for terms and substitutions. Actually, 
this removal is not without harm. Consider for example the case of relating the
empty substitution $\Gamma\vdash \epsilonu \Rightarrow \emptyu $
with a semantic substitution $\model\sigma$ between semantic contexts
$\model\Gamma$ and $\model\Delta$. We would like to assert that $\model\sigma$
equals the empty semantic substitution $\model\epsilon$, but this is not
possible because typechecking requires that $\model\Delta$ is 
the empty semantic context. This is precisely what ensured the removed hypothesis
$\syn\cdot \sim \model\Delta$.
Our way out here is to state that $\model\sigma$ relates to the empty
substitution if the target semantic context $\model\Delta$ is empty, and, acknowledging
this equality, if $\model\sigma$ is
the empty substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

@@

\subsection{Recursion}
Recursion consists in building a morphism from the syntactic model to the
semantic one.
The image of a syntactic context is going to be a semantic context which relates
to it (which is unique thanks to right uniqueness of the relation), and
similarly for types and terms.
% We construct it as the semantic
% Induction on the typing judgments shows that any QIIT morphism from the
% syntactic model $S$  to a semantic model $M$ sends well-typed syntax on a related
% semantic counter-part.
Thus, as a first step, we construct by induction on well-typed judgments the
related semantic counter-parts:
  \begin{alignat*}{5}
     & \Sigma\Con{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    \sum_{\model{\Gamma}} \wf{\Gamma} \sim \model{\Gamma}
    \\
     & \Sigma\Ty{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ({\model{A}:\model{\Ty}\model\Gamma})\times ( \wf{A} \sim \model{A})
    \\
     & \Sigma\Tm{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{t} : \Gamma \vdash t \in A ) \ra
    ({\model{t}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{t} \sim \model{t})
    \\
     & \Sigma\Var{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{A} : \Gamma \vdash A ) \ra
    ( \wf{A} \sim \model{A}) \ra
    \\ & && \qquad
    (\wf{x} : \Gamma \vdash x \invar A ) \ra
    ({\model{x}:\model{\Tm}\model\Gamma}\model{A})\times ( \wf{x} \sim \model{x})
    \\
     & \Sigma\Sub{\sim} && :
    (\wf{\Gamma} : \Gamma \vdash ) \ra
    (\wf{\Gamma}\sim \model\Gamma) \ra
    (\wf{\Delta} : \Delta \vdash  ) \ra
    ( \wf{\Delta} \sim \model{\Delta}) \ra
    \\ & && \qquad
    (\wf{\sigma} : \Gamma \vdash \sigma \Rightarrow \Delta ) \ra
    ({\model{\sigma}:\model{\Sub}\model\Gamma}\model{\Delta})\times ( \wf{\sigma} \sim \model{\sigma})
  \end{alignat*}
  Right uniqueness of the relation is used in this induction.
